%!TEX root = ../main.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Links:
%
% Difficulty:
% Companies: 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Counts the items on the containers}
\label{ch:items_in_containers_amazon}
\section*{Introduction}
Imagine you are the owner of a succesfully online store. 
You would like to be able to query what is the number of items you still have in the warehouse.
The problem is that you cannot just walk into the warehouse and count the items as they are stored in closed containers. 
Thankfully, the warehouse is equipped with sensors and it is able to produce a string representing 
the state of the warehouse and single containers. 
The problem described in this chapter investigates how we can write an algorithm
that takes such string (the state of all the containers in the warehouse) and is able to answer queries on the number of elements that are present 
in some portions of the warehouse itself. 

This problem has been reported to be asked during Amazon interviews and it is considered a medium difficulty problem.
We will investigate two solutions:
\begin{itemize}
	\item brute-force, that is going to have a relatively straightforward logic (blindly count the items in the string)
	 and be easy to code (in Section \ref{}),
	\item a more sophisticated one with optimal asymptotic complexity where the input string is preprocessed so that queries can be answered faster.
\end{itemize}



\section{Problem statement}
\begin{exercise}
	You are given a string $s$ representing the current state of a warehouse. 
	The string contains only two kinds of characters: 
	\begin{description}
		\item[\bsq{\textbf{\texttt{*}}}(ASCII 42)]: representing an item 
		\item[\bsq{\textbf{\texttt{|}}}(ASCII 124)]: representing the boundaries of a container.
	\end{description}
	
	A container is a closed space within the warehouse and it is represented in $s$ by a pair of  \bsq{\texttt{|}}.
	Items within a container $c$ are represented as \bsq{\texttt{*}} appearing withing the two \bsq{\texttt{|}} defining $c$.
	You are also given an array of pairs $Q = \{(s_0, e_0),(s_1, e_2),\ldots,(s_{n-1}, e_{e-1}) : 0 \leq s_i \leq e_i \leq |s|\}$, 
	the queries, where each pair in $Q$ identifies a substring in $s$.
	
	Your task is to write a function that returns an array $A$ of length $n$ 
	(containing the answers for each of the queries in $Q$)
	where each element $A_i$ is the number of items contained in all the \textbf{closed} compartments between $(s_i, e_i)$.
	

\begin{example}
	\hfill \\
	Given \texttt{s = \bsq{|**|*|*}} and $Q = \{(0,4),(0,5)\}$ the function returns $A=\{2,3\}$.
	$s$ has a total of $2$ closed containers the first with $2$ and $1$ item inside respectively.
	
	The first query asks you to find the number of elements in the substring $s[0,4]=$ \texttt{\bsq{|**|*}} 
	where three items are represented but only two are within a closed container (the first two).
	
	The second query refers to the substring $s[0,5]=$ \texttt{\bsq{|**|*|}}. The items are the same as in the previous query but this time
	all of them are in closed containers.
	
\end{example}

\begin{example}
	\hfill \\
	Given \texttt{s = \bsq{*|*|}} and $Q = \{(0,2),(1,3)\}$ the function returns $A=\{0,1\}$.
	$s$ has a total of two items and only $1$ closed container containing only a single item.

	The first query refers to the substring $s[0,2]=$ \texttt{\bsq{*|*}}. No closed container are represented in such substring thus the answer in this case must be $0$.
	However, the second question refers to  $s[1,3]=$ \texttt{\bsq{|*|}} where we can see we have a valid container. We can therefore counts the elements in it.
\end{example}

\end{exercise}
\section{Clarification Questions}

\begin{QandA}
	\item 
	\begin{answered}
		\textit{}
	\end{answered}
	
\end{QandA}

\section{Discussion}
\label{items_in_containers_amazon:sec:discussion}


\subsection{Brute-force}
\label{items_in_containers_amazon:sec:bruteforce}

\lstinputlisting[language=c++, caption={Sample Caption},label=list:items_in_containers_amazon]{sources/items_in_containers_amazon/items_in_containers_amazon_solution1.cpp}

